---
title: "Accessors for SummarizedExperiments"
author: "Martijn van Attekum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Accessors for SummarizedExperiments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The SummarizedExperiment's native col- and rowdata accessor functions 
(colData and rowData) have a number of non-tidyverse compliances:

* They return or require a DFrame object instead of a tibble
* When setting col-/rowdata they cannot be easily used in a pipe

cleanse offers pipe-friendly accessor functions that work with tibbles

```{r message=FALSE}
library(cleanse)
```
```{r}
data(seq_se)
# Get the coldata as a tibble
get_col_data(seq_se)

# generate new rowdata and create a new se with the rowdata set
new_rd <- get_row_data(seq_se) %>% 
  tidyr::unite(gene_group_name, c(gene_group, gene_name))
se_new_rowdata <- set_row_data(seq_se, new_rd)
```

A frequently used piece of data from an se is the options contained in its
coldata. cleanse offers an accessor function for this purpose:
```{r}
options_from_coldata(seq_se, "site")
```
```{r}
sessionInfo()
```
