---
title: "Arithmetic operations on SummarizedExperiments"
author: "Martijn van Attekum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Arithmetic operations on SummarizedExperiments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Occasionally, we want to perform arithmetic operations on the values contained in
2 se's, for instance when subtracting control values from treated values.  

If we want to subtract the values from T=0 from the values at T=4, this can
be achieved with a combination of ```select``` and ```-```. Take into account that columns are
subtracted by position and we have to ensure that the columns match up. 
The ```arrange``` function might be helpful to ensure this. (as shown below, not strictly
necessary for the example se though)
```{r}
library(dplyr)
library(SummarizedExperiment)
library(cleanse)
seq_se_t0 <- seq_se %>% select(time == 0) %>% arrange(col, patient, site, treatment)
seq_se_t4 <- seq_se %>% select(time == 4) %>% arrange(col, patient, site, treatment)
seq_se_diff <- seq_se_t4 - seq_se_t0
```
Similarly, functions for ```+```, ```/```, and ```*```, can be used.
 
