---
title: "Arithmetic operations on SummarizedExperiments"
author: "Martijn van Attekum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Arithmetic operations on SummarizedExperiments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Occasionally, we want to perform arithmetic operations on the values contained in
2 se's, for instance when subtracting control values from treated values.  

If we want to subtract the values from time == 0 from the values at time == 4, this can
be achieved with a combination of `filter` and `-`.  
Arithmetic operations require the se's to have the same ColData and rowData. 
To ensure the colData is the same, we have to:

* remove the time information from colData using the `select` function
* rearrange the data by patient, then site, then treatment (although not strictly
necessary for the example se)

```{r message=FALSE}
library(dplyr)
library(SummarizedExperiment)
library(cleanse)
```
```{r}
data(seq_se)
seq_se_t0 <- seq_se %>% filter(col, time == 0) %>% 
  select(col, -time) %>% 
  arrange(col, patient, site, treatment)
assay(seq_se_t0)[1:5, 1:5]
seq_se_t4 <- seq_se %>% filter(col, time == 4) %>% 
  select(col, -time) %>% 
  arrange(col, patient, site, treatment)
assay(seq_se_t4)[1:5, 1:5]
seq_se_diff <- seq_se_t4 - seq_se_t0
assay(seq_se_diff)[1:5, 1:5]
```
Similarly, functions for ```+```, ```/```, and ```*```, can be used.

In addition, a method to round the values from all assays is provided:
```{r}
seq_se_rounded <- seq_se %>% round(3)
assay(seq_se)[1:5, 1:5]
assay(seq_se_rounded)[1:5, 1:5]
```

```{r}
sessionInfo()
```
