---
title: "Manipulating SummarizedExperiments with dplyr functions"
author: "Martijn van Attekum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Manipulating SummarizedExperiments with dplyr functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The SummarizedExperiment (se) class offers a useful way to store multiple row and column
data along with the values from an experiment and is widely used in computational biology.
Although subsetting se's is possible with base R notation (ie using `[]`), se's lack methods
for dplyr functions such as `filter` and `select`, which excludes se's from easily being used in pipes.
This package offers methods for the dplyr functions and will automatically dispatch
your se to the relevant function.

The package contains an example se called `seq_se`. The example contains
dummy data of expression and copy number values for 10 genes and 48 different conditions.
To get an overview of the different options available in the se, use print_options:
```{r}
library(dplyr)
library(SummarizedExperiment)
library(cleanse)
print_options(seq_se)
```
  
As you can see, seq_se contains expression and copy_number data for samples taken from
3 different sites, which were treated with either treatment A or B for 0 or 4 hours for 4
different patients. The genes that were sequenced are from 3 different gene groups:
IL, NOTCH, and TLR

## Using dplyr functions
The `filter` and `select` functions from the dplyr package allow you to select rows and
columns respectively from a data frame and have been designed such that they integrate
into pipes. This functionalty has here been extended to the SummarizedExperiment.

```{r}
seq_se %>% filter(gene_group == "IL")
seq_se %>% select(treatment == "B", site %in% c("brain", "skin"))
```

Note that the selection criteria for the `select` function follow the notation of dplyr's
`mutate` function, where the variable names are based on the names from the se's attached colData.

The `arrange`, sample, and `mutate` function in dplyr act rowwise. As se's contain both row-
and colData, an extra argument is now supplied to have the action performed on either
the `row` or the `col`.

For instance, we can use the `arrange` function to consecutively sort a data frame by rows. 
In this case, we first arrange by gene_group and next by gene_name.
```{r}
seq_se %>% arrange(row, gene_group, gene_name)
```

The extended sample_n and sample_frac functions filters n random rows/cols 
or a random fraction of the rows/cols respectively
```{r}
seq_se %>% sample_n(col, size = 4, replace = TRUE)
seq_se %>% sample_frac(row, size = .5)
```

Lastly, mutate is normally used to add or change columns to a data frame based on 
existing columns. Again, the new variables can be added to the colData or the 
rowData of the SummarizedExperiment. If we for instance want to change the time 
for the samples from hours to minutes, we can do
```{r}
seq_se %>% mutate(col, time = (time * 60))
```
or to combine the gene groups and gene names to one new variable in the rowData
```{r}
seq_se %>% mutate(row, group_and_name = paste(gene_group, gene_name, sep = "_"))
```
