---
title: "Manipulating SummarizedExperiments with dplyr functions"
author: "Martijn van Attekum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Manipulating SummarizedExperiments with dplyr functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The SummarizedExperiment (se) class offers a useful way to store multiple row and column
data along with the values from an experiment and is widely used in computational biology.
Although subsetting se's is possible with base R notation (ie using `[]`), se's lack methods
for dplyr functions such as `filter` and `select`, which excludes se's from easily being used in pipes.
This package offers methods for the dplyr functions and will automatically dispatch
your se to the relevant function.

As se's contain a data frame for both the rowData and colData, a major difference using
these functions in cleanse, is that we need to **specify whether we apply our function to
the row or the col** of the se. Cleanse will then take care of updating the se.

The package contains an example se called `seq_se`. The example contains
dummy data of expression and copy number values for 10 genes and 48 different conditions.
To get an overview of the different options available in the se, use print_options:
```{r message=FALSE}
library(dplyr)
library(SummarizedExperiment)
library(cleanse)
```
```{r}
print_options(seq_se)
```
  
As you can see, seq_se contains expression and copy_number data for samples taken from
3 different sites, which were treated with either treatment A or B for 0 or 4 hours for 4
different patients. The genes that were sequenced are from 3 different gene groups:
IL, NOTCH, and TLR. Each of the rowData / colData columns (eg patient, site, gene_group)
is refered to as a **variable** henceforth.

## Using dplyr functions
The `filter` function subsets rows / cols from the se based on conditional filtering
of a variable contained in either rowData or colData respectively.
Experiment values in the assay are dropped concurrently with the update of the colData/rowData.
```{r}
genes_subset_se <- seq_se %>% filter(row, gene_group == "IL")
print_options(genes_subset_se)  #note the change in available gene_groups
dim(seq_se)
dim(genes_subset_se)

sample_subset_se <- seq_se %>% filter(col, treatment == "B", site %in% c("brain", "skin"))
print_options(sample_subset_se) #note the change in available treatments and sites
dim(seq_se)
dim(sample_subset_se)
```

`select` can be used to select variables from the attached rowData or colData, 
and `rename` will rename these variables. This will not remove any of the experiment's assay values.
```{r}
# remove the time variable after filtering for time == 0
seq_se_min_time <- seq_se %>% filter(col, time == 0) %>% select(col, -time)
print_options(seq_se_min_time)  #note the time variable has disappeared from the colData

# rename the time variable after changing it to minutes
seq_se_ren_time <- seq_se %>% mutate(col, time = (time * 60)) %>% rename(col, time_mins = time)
print_options(seq_se_ren_time)  #note the time variable is now called time_mins
```

`arrange` is used to consecutively sort a data frame by a variable. In this case, 
we arrange the rows, first by gene_name and next by gene_group.
```{r}
seq_se_reordered <- seq_se %>% arrange(row, gene_name, gene_group)
rowData(seq_se)
rowData(seq_se_reordered)
```

The extended `sample_n` and `sample_frac` functions filter n random rows/cols 
or a random fraction respectively
```{r}
seq_se %>% sample_n(col, size = 4, replace = TRUE) #note the change in dim

seq_se %>% sample_frac(row, size = .5) #note the change in dim
```

Lastly, `mutate` adds or changes variables from the rowData / colData.  
If we for instance want to change the time for the samples from hours to minutes, 
we can do
```{r}
seq_se_mins <- seq_se %>% mutate(col, time = (time * 60))
seq_se$time
seq_se_mins$time
```
or to combine the gene groups and gene names to one new variable in the rowData
```{r}
seq_se_gene_comb <- seq_se %>% 
  mutate(row, group_and_name = paste(gene_group, gene_name, sep = "_"))
rowData(seq_se_gene_comb)$group_and_name
```
