---
title: "Writing SummarizedExperiments to a delimited text file"
author: "Martijn van Attekum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Writing SummarizedExperiments to a delimited text file}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 
Due to their unique ability to contain row- and colData, SummarizedExperiments (se's) 
can not be written to a delimited file using functions such as write_csv. 
A set of functions is provided to mimic the write_<format> functions of readr.
Currently, read_csv, read_tsv, and read_delim can be used.
To round the example `seq_se` and write it to a csv file, use
```{r message=FALSE}
library(dplyr)
library(SummarizedExperiment)
library(cleanse)
```
```{r eval=FALSE}
seq_se %>% round(3) %>% write_csv("out.csv", "expression")
```
When not providing a folder, the output is written to the current working directory.
The beginning of the out.csv file will look like:
```
EXPRESSION gene_group gene_name                            
   patient                           1      2      3      4 ...
      site                        skin   skin   skin   skin ...
      time                           0      0      0      0 ...
 treatment                           A      A      A      A ...
                   IL     IL1R1 15.053  2.617 20.093 12.464 ...
                   IL     IL1R2  4.876 16.154 19.834  6.419 ...
                  ...       ...    ...    ...    ...    ... ...
```